Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-10-09 paper1_jiturra-UB-149.Rmd
ajustes patrick 3

Over the past few decades, cross-national studies on social class and redistributive preferences have predominantly centered on individual class positions [@lindh_class_2020]. Recently, however, a growing body of research has begun to focus on how the class profile of social networks influences redistributive preferences [@lindh_missing_2021; @lee_consider_2023; @paskov_crossclass_2022]. Although prior studies have shown that class-based network segregation—defined as the concentration of similar-class network ties—can undermine attachment to society [@otero_lives_2022], there is still limited understanding of how being embedded in a homogeneously class-based network impacts attitudes toward inequality and redistribution [@otero_power_2023]. Contributing to this debate, this research first aims to investigate the individual-level relationship between class-based network segregation and redistributive preferences. In addition, it has been suggested that social classes tend to form segregated networks according to the homophily principle, understood as the higher chances of forming social ties with others with similar characteristics [@mcpherson_birds_2001;@otero_lives_2022]. Furthermore, cross-national comparisons indicate that class-based network stratification strengthens in societies with high economic inequality, which in turn drives stronger social distance between classes, particularly among the lower classes [@otero_differences_2023; @pichler_social_2009]. Against this backdrop, the present study seeks to contribute to understanding how cross-national differences in income inequality act as a moderating factor in the relationship between class-based network segregation and redistributive preferences.

While segregation may deepen existing class divisions in redistributive preferences [@paskov_crossclass_2022], little is known about how income inequality shapes this relationship. On the one hand, research on the class-attitude link suggests that income inequality is crucial for understanding how class relations—i.e., the social and political distance between classes—translate into redistributive demands, as it reflects the state of class struggle in contemporary capitalist societies [@edlund_democratic_2015; @curtis_how_2015]. A consistent finding is that the upper classes tend to be more supportive of redistribution in contexts of high inequality, whereas the already high redistributive demands of the lower classes remain relatively stable, regardless of inequality levels [@sachweh_why_2019; @dimick_altruistic_2017]. On the other hand, it is well-documented that income inequality affects social relations, reinforcing stratified access to social activities and widening the distance between classes. In such conditions, the upper classes tend to be more socially active and maintain more diverse networks, while the lower classes become increasingly inactive and segregated as inequality rises [@lancee_income_2012; @otero_differences_2023; @pichler_social_2009]. However, current research efforts have primarily examined the impact of income inequality moderates the relations of social class on either social networks or redistributive preferences. Thus, this study aims to address two key questions:

@@

Theoretically, studies on economic inequality, social class, and redistributive preferences show varying class reactions to rising inequality. Political economists suggest that high-income individuals are far from being monolithic in their redistributive preferences, motivated by concerns about inequality's adverse effects (e.g., crime), leading to altruistic support for redistribution [@dimick_models_2018; @dimick_altruistic_2017; @rueda_externalities_2016]. Conversely, the moral economy literature in sociology explains affluent differences through evaluations of distributive justice regarding resource allocation procedures [@liebig_sociology_2016]. At the same time, the literature on the role of income inequality on social cohesion has argued that unequal societies strengthen barriers to trustworthy social connections. Neckerman and Torche [-@neckerman_inequality_2007, p. 344] noted that marginalization is linked to reduced participation in social life, exacerbated by greater inequality. Also, rising economic inequality erodes social relations by reducing social trust, which underpins mutual understanding and collective solidarity [@kragten_income_2017]. Conversely, it is noted that societies with low inequality and comprehensive welfare states enhance civic and social participation, increasing generalized trust and reducing social conflict [@uslaner_inequality_2005]. Similarly, it has been argued that social participation can nurture higher chances for cross-class interactions and motivate more egalitarian attitudes, being upper-class citizens who support stronger collective solidarity through government redistribution in more cohesive communities [@yamamura_social_2012].

Empirically, it has been shown that affluent groups are more responsive, perceiving that rising inequality affects the overall opportunity structure and social mobility chances [@sachweh_why_2019]. Also, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of procedural justice [@kim_socioeconomic_2018]. In contrast, low-income individuals view ascribed characteristics as more crucial in constraining opportunities, irrespective of current income inequality [@sachweh_why_2019]. Similarly, @edlund_democratic_2015 argue that in unequal societies with residual welfare states, social class lacks significant political meaning, weakening distributive conflicts around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can be attributed to their greater awareness of the incentives and societal consequences of income inequality [@curtis_how_2015; @Svallfors2006]. At the same time, cross-national comparative evidence indicates that stratified access to social activities and networks is reinforced in unequal societies. For example, @pichler_social_2009 demonstrated that class differences in civic and family network participation are exacerbated in unequal societies. Similarly, @lancee_income_2012 found that income-based stratification in civic participation increases with inequality. Also, @otero_differences_2023 found that economic inequality enhances 
interdependence among cultural, economic, and social capital, which results in stronger stratified chances for cross-class interactions. Thus, in unequal societies, the upper classes can navigate diverse social settings while maintaining their distinct positions, whereas the lower classes remain marginalized and segregated due to others' choices [@otero_open_2021, p.24].

Against this backdrop, there are at least three emerging theoretical implications on the role of income inequality on class-based networks and redistributive preferences. The first implication is that it is plausible to assume that income inequality drives greater segregation, as the result of lower participation and social trust. However,  it is worth noticing that resource interdependency seems to exacerbate social exclusion for the lower classes, while the upper classes may hold more cohesive experiences that drive stable or even lower segregation in more unequal societies. A second implication is that the relationship between class and attitudes weakens in highly unequal societies, where the upper classes tend to show more support for redistribution, contrasting with the relatively stable preferences of the lower classes. In addition, it is also true that in egalitarian societies with more encompassing welfare states class-based distributive struggles tend to be more institutionalized (e.g. through unions or political parties). As a result, class positions are more closely implied in political conflict as the institutional framework allows them to better articulate their interests. A third implication is that as inequality increases and the class divide in redistributive preferences narrows, this may also be extended to the influence of class-based segregation on redistributive preferences. As the class-attitude link becomes mild in more unequal contexts, it is also expected that homogeneous class-based networks might not have the same strength where the role of class positions loses significance. Therefore, I expect income inequality to weaken the interaction between network homogeneity and social class. Specifically, I hypothesize that the conditional relationship between network homogeneity and social class stated at the individual level (*segregation hypothesis*) will be less pronounced in the context of higher income inequality (*mitigation hypothesis*). Given these considerations, the second hypothesis is as follows:

>>>>

Over the past few decades, cross-national studies on social class and redistributive preferences have predominantly centered on individual class positions [@lindh_class_2020]. Recently, however, a growing body of research has begun to focus on how the class profile of social networks influences redistributive preferences [@lindh_missing_2021; @lee_consider_2023; @paskov_crossclass_2022]. Although prior studies have shown that class-based network segregation—defined as the concentration of similar-class network ties—can undermine societal cohesion [@otero_lives_2022], there is still limited understanding of how being embedded in a homogeneously class-based network impacts attitudes toward inequality and redistribution [see @otero_power_2023]. Contributing to this debate, this research first aims to investigate the individual-level relationship between class-based network segregation and redistributive preferences. At the same time, it has been suggested that social classes tend to form segregated networks [see @mcpherson_birds_2001; @otero_lives_2022], and cross-national comparisons indicate that in more economically unequal societies, the link between social class and network composition strengthens, leading to greater segregation, particularly among the working class [see @otero_differences_2023; @lancee_income_2012].

Against this backdrop, the present study seeks to explore how cross-national differences in income inequality act as a moderating factor in the relationship between class-based network segregation and redistributive preferences. While segregation may deepen existing class divisions in redistributive preferences [@paskov_crossclass_2022], little is known about how income inequality shapes this relationship. On theone hand, research on the class-attitude link suggests that income inequality is crucial for understanding how class relations—i.e., the social and political distance between classes—translate into redistributive demands, as it reflects the state of class struggle in contemporary capitalist societies [@edlund_democratic_2015; @curtis_how_2015]. A consistent finding is that the upper classes tend to be more supportive of redistribution in contexts of high inequality, whereas the already high redistributive demands of the working class remain stable, regardless of inequality levels [@sachweh_why_2019; @dimick_altruistic_2017].  On the other hand, it is well-documented that income inequality affects social relations, reinforcing stratified access to social activities and widening the distance between classes. In such conditions, the upper classes tend to be more socially active and maintain more diverse networks, while the lower classes become increasingly inactive and segregated as inequality rises [@lancee_income_2012; @otero_differences_2023; @pichler_social_2009]. However, current research efforts have primarily examined the impact of income inequality either on social networks or on support for redistribution. Thus, a key question remains: to what extent does income inequality moderate the relationship between class-based network segregation and redistributive preferences?

Therefore, this study aims to address two key questions:

@@

Studies on economic inequality, social class, and redistributive preferences show varying class reactions to rising inequality. Political economists suggest that high-income individuals are far from being monolithic in their redistributive preferences, motivated by concerns about inequality's adverse effects (e.g., crime), leading to altruistic support for redistribution [@dimick_models_2018; @dimick_altruistic_2017; @rueda_externalities_2016]. Conversely, the moral economy literature in sociology explains affluent differences through evaluations of distributive justice regarding resource allocation procedures [@liebig_sociology_2016]. Furthermore, income inequality strengthens barriers to trustworthy social connections, affecting social relations. Neckerman and Torche [-@neckerman_inequality_2007, p. 344] noted that marginalization is linked to reduced participation in social life, exacerbated by greater inequality. Higher economic inequality also erodes social relations by reducing social trust, which underpins mutual understanding and solidarity [@kragten_income_2017]. Conversely, lower-income inequality and comprehensive welfare states enhance civic and social participation, increasing generalized trust and reducing social conflict [@uslaner_inequality_2005]. Additionally, @yamamura_social_2012 found that redistributive preferences increase with community-level social activity participation, with high-income citizens supporting collective solidarity through government redistribution.

Empirically, it has been shown that affluent groups are more responsive, perceiving that rising inequality affects the overall opportunity structure and social mobility chances [@sachweh_why_2019]. Also, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of procedural justice [@kim_socioeconomic_2018]. In contrast, low-income individuals view ascribed characteristics as more crucial in constraining opportunities, irrespective of current income inequality [@sachweh_why_2019]. Similarly, @edlund_democratic_2015 argue that in unequal societies with residual welfare states, social class lacks significant political meaning, weakening distributive conflicts around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can be attributed to their greater awareness of the incentives and societal consequences of income inequality [@curtis_how_2015; @Svallfors2006]. At the same time, comparative studies indicate that stratified access to social activities and networks is reinforced in unequal societies. For example, @pichler_social_2009 demonstrated that class differences in civic and family network participation are exacerbated in unequal societies. Similarly, @lancee_income_2012 found that income-based stratification in civic participation increases with inequality. Also, @otero_differences_2023 found that economic inequality enhances stratified access to network diversity, amplifying interdependence among cultural, economic, and social capital. Thus, the upper classes navigate diverse social settings while maintaining their distinct positions, whereas the lower classes remain marginalized and segregated due to others' choices [@otero_open_2021, p.24].

Against this backdrop, one theoretical implication is that income inequality drives greater segregation. However, resource interdependency seems to exacerbate social exclusion for the lower classes, while the upper classes may experience less segregation. Another inference is that the relationship between class and attitudes weakens in highly unequal societies, where the upper classes tend to show more support for redistribution, contrasting with the relatively stable preferences of the lower classes. In addition, it is also true that in egalitarian societies with more encompassing welfare states class-based distributive struggles tend to be more institutionalized (e.g. through unions or political parties). As a result, class positions are more closely implied in political conflict as the institutional framework allows them to better articulate their interests.  A derived implication is that as inequality increases and the class divide in redistributive preferences narrows, this may also be extended to the influence of class-based segregation on redistributive preferences. Therefore, I expect income inequality to weaken the interaction between network homogeneity and social class. Specifically, I hypothesize that the conditional relationship between network homogeneity and social class (*segregation hypothesis*) will be less pronounced in the context of higher income inequality (*mitigation hypothesis*). Given these considerations, the second hypothesis is as follows:
